Some lessons after the formalization of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-05-04 JSL/lessons.tex
Preliminary analysis of instances; a few words on proof assistants
types and sorts in Isabelle's M

  Note: This is in direct contradiction with “code fever”
  (\ref{sec:beware-code-fever} below).
\end{enumerate}

\subsection{Read the fine print}
\begin{enumerate}
\item Foundations of your proof assistant.
\end{enumerate}

@@

\subsection{Beware of the “Code fever”}\label{sec:beware-code-fever}
\begin{enumerate}
\item “We know that doing math is fun---formalization is like DRUGS”
\item Feeling of accomplishment after seeing your writings
  validated beyond reasonable doubt.
\item One easily forgets about the “Power of the Board.”
\end{enumerate}

\subsection{The Devil's on the shortcuts}
\begin{enumerate}
\item
  Our proofs of the "definition of forces" (and many
  consequences) and of the lemma for "forcing a value" of function
  depend on the countability of the ground model. 
\item
  Density arguments (look for "TODO", "general versions").
\end{enumerate}

>>>>

  Note: This is in direct contradiction with “code fever”
  (\ref{sec:beware-code-fever} below).
\item Read the fine print: Foundations of your proof assistant.
\end{enumerate}

@@

\subsection{Beware of the “Code fever”}\label{sec:beware-code-fever}
\begin{enumerate}
\item “We know that doing math is fun---formalization is like DRUGS”
\item Feeling of accomplishment after seeing your writings
  validated beyond reasonable doubt (v.g. cofinality).

\item One easily forgets about the “Power of the Board.”
\end{enumerate}

\subsection{The Devil's on the shortcuts}
\begin{enumerate}
\item
  Our proofs of the “definition of forces” (and many
  consequences) and of the lemma for “forcing a value” of function
  depend on the countability of the ground model. 
\item
  Density arguments (look for “TODO”, “general versions”).
\end{enumerate}

### 2022-05-04 JSL/proof_assists_isabelle.tex
types and sorts in Isabelle's M

Several of the more established assistants (HOL Light, Coq, Isabelle)
are programmed in some variant of the ML language (which was
originally designed for this purpose); the newer Lean, on the other
hand, was originally conceived as a standalone functional programming
language with all the features of a standard assistant.

@@

>>>>

Several of the more established assistants (HOL Light, Coq, Isabelle)
are programmed in some variant of the ML language (which was
designed for this purpose); the newer Lean, on the other
hand, was originally conceived as a standalone functional programming
language with all the features of a standard assistant.

@@

The logical foundation of Isabelle is an intuitionistic fragment of
higher-order logic (or simple type theory) called $\Meta$;
original version was described in \cite{Paulson1989}, and the addition
of  sorts appears in \cite{Nipkow-LF-91}.

The only predefined type is $\prop$ (“propositions”); new base types
can be postulated when defining objects logics. Types of higher order can be
assembled using the function space constructor $\fun$.

The type of propositions $\prop$ equipped with a binary operation
$\implies$ (“meta-implication”) that is used to represent the object
logic rules. As an example, the axiomatization of first-order logic
postulates a type $\tyo$ of booleans, and Modus Ponens
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#IFOL.mp|axiom
is written as
\[
  [P\limp Q] \implies ([P] \implies [Q]).
\]
The square brackets represent an injection from $\tyo$ into
$\prop$. % ($[P]$ can be read as “$P$ holds”)
A consequence of this representation is that every
formula of the object logic appears as atomic for $\Meta$.

%% Quantification is handled in the meta-level using a functional $\ALL$
%% with polymorphic type $(\alpha \fun \prop) \fun \prop$. 
Types in Isabelle are organized into \emph{classes} and \emph{sorts};
for ease of exposition, we will omit the former.  The axiomatization
of first-order logic postulates a sort $\type{term}$ (of
“individuals”, or elements of a first-order universe of discourse) and
stipulates that every further type variable $\alpha$ must be of that
sort. In particular, Isabelle/ZF only postulates one new type $\tyi$
(“sets”) of sort $\type{term}$. Hence, from the type of the universal
quantifier functional $\forall :: (\alpha \fun \tyo) \fun \tyo$ it
follows that it may only be applied to predicates with a variable of
type $\tyi$. This ensures that the object logic is effectively
first-order.

Paulson  \cite{Paulson1989} carried out a proof that the encoding
$\Meta_{\mathrm{IFOL}}$ of
intuitionistic first-order logic without equality in the original $\Meta$ is
conservative (there is a correspondence between provable $\phi$ in
IFOL and provable $[\phi]$ in $\Meta_{\mathrm{IFOL}}$) by putting
$\Meta_{\mathrm{IFOL}}$ proofs in \emph{expanded normal form}
\cite{MR0387024}. Passing to classical logic does not presents
difficulties, but the treatment of meta-equality differs between the
original and the present incarnation of $\Meta$; details are
exhaustively expounded in the recent formalization by Nipkow and
Roßkof \cite{10.1007/978-3-030-79876-5_6}.
